===============================================================================================================

# Peer Assignment 1

===============================================================================================================


## Loading and preprocessing the data

It is assumed that the "activity" file is downloaded, unziped and placed in the working directory.


```{r, echo = TRUE}
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
str(data)


data$date <- as.Date(data$date)
str(data)

sum(is.na(data$steps))
```


## Mean total number of steps per day

Since we can ignore the NA's for this part, NA's are temporarily excluded from this analysis.

data_rm doesn't contain any NA's
```{r, echo = TRUE}
data_rm<-data[which(!is.na(data$steps)),]
```

The number of steps taken is measured in 5-minute intervals, In order to compute the total number of steps taken for each day we will aggregate the data by day.
```{r, echo = TRUE}
perday<-tapply(data_rm$steps, data_rm$date, sum)
```

Make a histogram of the total number of steps taken each day and display its mean & median.
```{r, echo = TRUE}
hist(perday,10, main = "Total number of steps taken per day", xlab = "")

mean(perday)

median(perday)
```

## Average daily activity pattern


Then aggregate the dataset by the intervals. A per interval array is created for this reason and a time series plot will also be created.
```{r, echo = TRUE}
dailyact<-tapply(data_rm$steps, data_rm$interval, mean)
plot(y = dailyact, x = names(dailyact), type = "l", xlab = "5-Minute-Interval", 
     main = "Daily Activity Pattern", ylab = "Average number of steps")
```

Display the interval with the maximum average number of steps throughout the days is 835 with 206.1698 steps.
```{r, echo = TRUE}
dailyact[dailyact==max(dailyact)]
```

## Replacing missing values
```{r, echo = TRUE}
sum(is.na(data$steps))
```

```{r, echo = TRUE}
sum(is.na(data))
```

In the original data set, the NA's are replaced with the mean of the 5 minute interval to create a new dataset.
```{r, echo = TRUE}
act_new <- data
act_new[which(is.na(act_new$steps)),1]<-
  dailyact[as.character(act_new[which(is.na(act_new$steps)),3])]
```

Check for missing values again
```{r, echo = TRUE}
sum(is.na(act_new))
```

Make the histogram again, that we made in the first part of the analysis, in order to see if there is a change.
```{r, echo = TRUE}
perday_new<-tapply(act_new$steps, act_new$date, sum)
```

```{r, echo = TRUE}
par(mfrow=c(1,2))
hist(perday,10, main = "Total number of steps taken per day", xlab = "Steps"
     , ylim =c(0, 25))
abline(v = median(perday), col = 4, lwd = 4)
hist(perday_new,10, main = "Total number of steps taken per day  
     (missing values replaced with mean of interval)", xlab = "Steps",
     ylim =c(0, 25))
abline(v = median(perday_new), col = 4, lwd = 4)
```

We now calculate the median and the mean of the filled in dataset
```{r, echo = TRUE}
mean(perday_new)

median(perday_new)
```

The effect of inputting missing data is minimal.
```{r, echo = TRUE}
mean(perday_new)-mean(perday)

median(perday_new)-median(perday)
```

## Difference in activity patterns between weekdays and weekends

Create factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r, echo = TRUE}
act_new$wd<-weekdays(act_new$date)
act_new$fwd<- as.factor(c("weekend", "weekday"))
act_new[act_new$wd == "Sunday" | act_new$wd == "Saturday" ,5]<- factor("weekend")
act_new[!(act_new$wd == "Sunday" | act_new$wd == "Saturday"),5 ]<- factor("weekday")
```

Create two aggregated arrays for the total number of steps taken per 5-minyute time interval for weekdays and weekends, and make a graph in order to compare it there is a difference.
```{r, echo = TRUE}
act_new_we <- subset(act_new, fwd == "weekend") 
act_new_wd <- subset(act_new, fwd == "weekday") 
dailyact_we<-tapply(act_new_we$steps, act_new_we$interval, mean)
dailyact_wd<-tapply(act_new_wd$steps, act_new_wd$interval, mean)
```

```{r scatterplot, fig.height=10}
par(mfrow=c(2,1))
plot(y = dailyact_wd, x = names(dailyact_wd), type = "l", xlab = "5-Minute Interval", 
     main = "Daily Activity Pattern on Weekdays", ylab = "Average number of steps", 
     ylim =c(0, 250))
plot(y = dailyact_we, x = names(dailyact_we), type = "l", xlab = "5-Minute Interval", 
     main = "Daily Activity Pattern on Weekends", ylab = "Average number of steps", 
     ylim =c(0, 250))
```

From the two graphs, we can see that the distribution throughout the day is different. An interesting finding is that there is a huge amount of steps taken on weekdays, possibly while going to work or working out, which does not appear on Saturdays or Sundays. 
